---
title: Market Basket Analysis
---
Market Basket Analysis (also known as affinity analysis or association rules mining) is an unsupervised machine learning technique which applies an algorithm (apriori algorithm) to identify association rules in datasets.

Whereas this algorithm is usually applied in a business setting, here we are using it to identify associations among our key variables of interest (clinical manifestation, vaccine period, serotype group, and age group).

We'll start with loading in our packages:

```{r}
library(tidyverse)
library(knitr)
library(ggplot2)
library(lubridate)
library(arules)
library(arulesViz)
library(plyr)
library(RColorBrewer)
library(plotly)
```

Next, we load in and clean our dataset:

```{r}
nepal.gps = read.csv("nepal_gps.csv")

# add list for all serotype groups
pcv10_serotypes <- c('1', '4', '5', '6B', '7F', '9V', '14', '18C', '19F', '23F') 
pcv13_serotypes <- c('3', '6A', '19A')
pcv15_serotypes <- c('22F', '33F')
pcv20_serotypes <- c('8', '10A', '11A', '12F', '15B')
untypable = c("ALTERNATIVE_ALIB_NT","COVERAGE TOO LOW", "SWISS_NT", "UNTYPABLE")

# categorize serotypes by "PCV10 Serotypes," "Other PCV Serotypes" (including additional PCV13, PCV15, and PCV20 serotypes), and "Non-PCV Serotypes" (those not targetted by any vaccine we looked at). There were also some cases where the serotype was not specified/untypable, which were grouped as "UNTYPABLE" and excluded from this analysis
nepal.gps$Serotype_Group <- ifelse(nepal.gps$In_silico_serotype %in% pcv10_serotypes, "PCV10 Serotypes",
                            ifelse(nepal.gps$In_silico_serotype %in% pcv13_serotypes, "Other PCV Serotypes",
                            ifelse(nepal.gps$In_silico_serotype %in% pcv15_serotypes, "Other PCV Serotypes",
                            ifelse(nepal.gps$In_silico_serotype %in% pcv20_serotypes, "Other PCV Serotypes",
                            ifelse(nepal.gps$In_silico_serotype %in% untypable, "UNTYPABLE",
                            "Non-PCV Serotypes")))))

# making age groups: 0-2, 3-5, and 6-14. All age data fell within this range, with many missing values. After reviewing literature we found that infants (age 0-2) were the most vaccinated group, and were highly covered by the vaccine. Children in Nepal do not go to school until age 6, so 6-14 represents school-age children. We also added a "No Age Data" category as we did not want to exclude all cases with missing age data.
age_0_2 <- c('1', '2') 
age_3_5 <- c('3', '4', '5')
age_6_plus <-c(6:14)
nepal.gps$Age_years <- ifelse(nepal.gps$Age_years %in% age_0_2, "Age 0-2",
                            ifelse(nepal.gps$Age_years %in% age_3_5, "Age 3-5",
                                   ifelse(nepal.gps$Age_years %in% age_6_plus, "Age 6-14",
                            "No Age Data")))

# removing all cases where the serotype was not specified, as we are looking at associations with each serotype group
nepal.gps <- nepal.gps[nepal.gps$Serotype_Group != "UNTYPABLE", ]

# post-vaccine periods are labeled at 1-year, 2-year, and 3-year follow-ups: we grouped them together here due to sparse data
post_vac <- c("POSTPCV10-1YR", "POSTPCV10-2YR", "POSTPCV10-3YR")

# editing vaccine period variable to be only Pre- or Post-Vaccine
nepal.gps$Vaccine_period <- ifelse(nepal.gps$Vaccine_period %in% post_vac, "Post-PCV10", "Pre-PCV10")

# selected only our variables of interest: removed cot, cat, fol (genes: negative or positive) and penicillin, gpsc, wgs_..., macrolide, etc: only focusing on age group, clinical manifestation, vaccine period, and serotype group. with more data, we may have considered looking at these variables as well, but to answer our research question we decided to focus on just four variables:
nepal.gps <- nepal.gps %>% 
  select(Age_years, Clinical_manifestation, Vaccine_period, Serotype_Group)

# renaming variable names for clarity and consistency for plotting
nepal.gps$Clinical_manifestation <- ifelse(nepal.gps$Clinical_manifestation == "UNSPECIFIED IPD", "Unspecified IPD", tools::toTitleCase(tolower(nepal.gps$Clinical_manifestation)))

# writing new data to a csv file: nepal.gps.csv
write.csv(nepal.gps, "nepal.gps.csv", row.names = FALSE)
```

Next, we create a transactions object for the arules package to use: using 'basket' format so every line represents a 'transaction', not a single item.

```{r}
tr<-read.transactions("nepal.gps.csv", format= 'basket', sep= ',')
```

Next, we print a description of the transaction: we see that "Carriage" was the most frequent item. 

```{r}
print('Description of the transactions')
summary(tr)
```

To see what items occur together most frequently, we create an item frequency plot:

```{r}
itemFrequencyPlot(tr,topN=25,type="absolute",col=brewer.pal(8,'Pastel2'), main="Item Frequency Plot")
```
Next we make a relative frequency plot:
 
```{r}
itemFrequencyPlot(tr,topN=20,type="relative",col=brewer.pal(8,'Pastel2'),main="Relative Frequency Plot")
```
## Create Rules

We use the apriori algorithm from the arules package to look for itemsets and find support for rules. 

We pass supp=0.01 and conf=0.8 to return all the rules have a support of at least 1% and confidence of at least 80%. 

We sort the rules by decreasing confidence. 

Here are the rules matching these criteria:

```{r}
rules <- apriori(tr, parameter = list(supp=0.01, conf=0.8))
rules <- sort(rules, by='confidence', decreasing = TRUE)
summary(rules)
```
We have 17 rules based on our criteria. We next inspect the top 10 rules according to these parameters (support=0.01, confidence=0.8).

```{r}
inspect(rules[1:10])
topRules <- rules[1:10]
```

We can plot these 17 rules by their lift, confidence, and support data:

```{r}
plot(rules)
```
Adding color to show the order of the rules:

```{r}
plot(rules, method = "two-key plot")
```

A network graph is a good way to visualize what these rules look like. We see two distinct clusters: one set of rules associated with the Pre-Vaccine period, and one set associated with the Post-Vaccine period. The top 5 rules (sorted by confidence) are all associated with the Pre-Vaccine period.

```{r}
plot(topRules, method="graph")
```
We can look at the rules using an interactive map as well.

```{r}
plot(topRules, method="graph", engine = 'interactive')
```

Another visualization:

```{r}
plot(topRules, method = "grouped")
```

One of the easier ways to visualize these rules is using an html widget, where the rules are ordered by confidence, sized by support, and colored by lift. 

```{r}
plot(topRules, method = "graph",  engine = "htmlwidget")
```

To create a figure for our poster and presentation, we change the color scheme

```{r}
# create the base plot
p <- plot(topRules, method = "graph", engine = "htmlwidget")

# create a list of the lift values for rules 1-10 (to create a gradient color scale)
lift_values <- c(
  1.755556,
  1.755556,
  1.755556,
  2.972326,
  2.918284,
  1.370537,
  1.339656,
  1.333718,
  1.329564,
  2.743685
)

# create a teal gradient from light to dark
teal_palette <- colorRampPalette(c("#b2d8d8", "#007a7a"))

# generate colors mapped to lift values (sorted low to high)
ranked_colors <- teal_palette(length(lift_values))[rank(lift_values)]

# final color list, aligned with our lift values
rule_teal_colors <- ranked_colors

# variables for the labels and the shapes: circles are rules, boxes are items
node_labels <- p$x$nodes$label
node_shapes <- p$x$nodes$shape

color_list <- vector("list", length(node_labels))
rule_index <- 1

# loop through node labels to assign colors
for (i in seq_along(node_labels)) {
  # if circle: it is a rule
  if (node_shapes[i] == "circle") {
    # rule nodes are assigned the correct color based on their lift
    color_list[[i]] <- list(
      background = rule_teal_colors[rule_index],
      border = "#00356B",
      highlight = rule_teal_colors[rule_index]
    )
    rule_index <- rule_index + 1
  # if not circle, the node is a box (meaning it an item): make all items gray
  } else {
    # assign color to be gray
    color_list[[i]] <- list(
      background = "#d3d3d3",
      border = "#aaaaaa",
      highlight = "#d3d3d3"
    )
  }
}

# extra code for styling purposes
p$x$nodes$color <- color_list
p$x$nodes$font <- lapply(seq_len(nrow(p$x$nodes)), function(i) {
  list(color = "#000000", size = 16, face = "bold")
})
p$x$options$nodes <- list(font = list(color = "#000000", size = 16, face = "bold"))
p$x$nodes$size <- rep(25, nrow(p$x$nodes))

# next, we change some of the physics configurations: adding in this extra code makes it more difficult to move our nodes around (moving one moves all the rest), which made it difficult to move the boxes close enough to create a final product that was readable on our poster/in our presentation. here we tried to make it easier to move things around:
p$x$options$physics <- list(
  enabled = TRUE,
  stabilization = list(
    enabled = TRUE,
    iterations = 1000,
    updateInterval = 25,
    onlyDynamicEdges = FALSE,
    fit = TRUE
  ),
  barnesHut = list(
    gravitationalConstant = -2000,
    springConstant = 0.0001,
    damping = 0.09
  )
)

# disable physics after our layout stabilizes
p$x$events <- list(
  stabilized = htmlwidgets::JS("
    function () {
      this.physics.disable();
    }
  ")
)

# visualize our final plot!
p
```

Last, we made a legend for this new color scheme: 

```{r}
library(ggplot2)

# vector of lift values
lift_values <- c(
  1.755556,
  1.755556,
  1.755556,
  2.972326,
  2.918284,
  1.370537,
  1.339656,
  1.333718,
  1.329564,
  2.743685
)

# color palette from light to dark
teal_palette <- colorRampPalette(c("#b2d8d8", "#007a7a"))

# number of bins: 100 makes for a smoother gradient
n_bins <- 100

# generate the data for our legend
legend_df <- data.frame(
  lift = seq(min(lift_values), max(lift_values), length.out = n_bins)
)

# plot to show horizontal color gradient
ggplot(legend_df, aes(x = lift, y = 1, fill = lift)) +
  geom_tile() +
  scale_fill_gradientn(
    colors = teal_palette(n_bins),
    name = "Lift"
  ) +
  scale_x_continuous(breaks = pretty(range(lift_values), n = 5)) +
  theme_minimal() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid   = element_blank(),
    axis.title.x = element_text(size = 15),
    legend.position = "none"  # hide the redundant legend for the legend
  ) +
  coord_fixed(ratio = 0.05) -> legend

# save image: wide & short for the final product
ggsave("legend.png", plot = legend, width = 4, height = 1.5)
```

The code above is for our most interpretable image, but there are several other ways to analyse our rules. We can make a matrix plot:

```{r}
plot(topRules, method = "matrix", engine = "3d", measure = "lift")
```