---
title: Market Basket Analysis Walkthrough
---
Market Basket Analysis, aka affinity analysis aka association rules mining is an unsupervised machine learning technique which applies an algorithm (apriori algorithm) to identify association rules in datasets.

We'll be applying this algorithm to identify associations in a dataset containing information about cases of perforative acute otitis media in children.

We'll start with loading in our packages

```{r}
library(tidyverse)
library(knitr)
library(ggplot2)
library(lubridate)
library(arules)
library(arulesViz)
library(plyr)
library(RColorBrewer)
library(plotly)
```

And our dataset

```{r}
nepal.gps = read.csv("nepal_gps.csv")

nepal.gps$macrolide_resistant <- ifelse(nepal.gps$ermB == "NEG" & nepal.gps$mefA == "NEG", "macrolide", "macrolide_resistant")

pcv10_serotypes <- c('1', '4', '5', '6B', '7F', '9V', '14', '18C', '19F', '23F') 
pcv13_serotypes <- c('3', '6A', '19A')
pcv15_serotypes <- c('22F', '33F')
pcv20_serotypes <- c('8', '10A', '11A', '12F', '15B')
untypable = c("ALTERNATIVE_ALIB_NT","COVERAGE TOO LOW", "SWISS_NT", "UNTYPABLE")
nepal.gps$Serotype_Group <- ifelse(nepal.gps$In_silico_serotype %in% pcv10_serotypes, "PCV10",
                            ifelse(nepal.gps$In_silico_serotype %in% pcv13_serotypes, "PCV13",
                            ifelse(nepal.gps$In_silico_serotype %in% pcv15_serotypes, "PCV15",
                            ifelse(nepal.gps$In_silico_serotype %in% pcv20_serotypes, "PCV20", 
                            ifelse(nepal.gps$In_silico_serotype %in% untypable, "UNTYPABLE",
                            "Other")))))

nepal.gps <- nepal.gps %>%
  select(Age_years, Clinical_manifestation, Penicillin, Vaccine_period, GPSC, WGS_PEN_SIR_Meningitis, Cot, folA_I100L, cat, macrolide_resistant, Serotype_Group)

cols_to_change1 <- c("Cot", "folA_I100L", "cat")

for (col in cols_to_change1) {
  nepal.gps[[col]] <- ifelse(nepal.gps[[col]] == "POS",
                             paste0("POS", col),
                             paste0("NEG", col))}

write.csv(nepal.gps, "nepal.gps.csv", row.names = FALSE)
```


```{r}
tr<-read.transactions("nepal.gps.csv", format= 'basket', sep= ',')
```

```{r}
print('Description of the transactions')
summary(tr)
```

Let's see what items occur most frequently:

```{r}
itemFrequencyPlot(tr,topN=25,type="absolute",col=brewer.pal(8,'Pastel2'), main="Nepal_gps rules")
```
a relative frequency plot
 
```{r}
itemFrequencyPlot(tr,topN=20,type="relative",col=brewer.pal(8,'Pastel2'),main="Relative frequency, Nepal_GPS")
```
## Create some rules

We use the Apriori algorithm from the arules package to look for itemsets and find support for rules

We pass supp=0.0001 and conf=0.8 to return all the rules have a support of at least 0.1% and confidence of at least 80%. 

We sort the rules by decreasing confidence. 

Here are the rules matching these criteria:

```{r}
rules <- apriori(tr, parameter = list(supp=0.01, conf=0.8))
rules <- sort(rules, by='confidence', decreasing = TRUE)
summary(rules)
```


We have 4093 rules, most are 4 or 5 items long. Let's inspect the top 10 rules according to these parameters (supp 0.001, conf =0.8).

```{r}
inspect(rules[1:10])
```

And plot these top 10 rules, or 20, or 50.

```{r}
topRules <- rules[1:10]
plot(rules)
```
now with more colors
```{r}
plot(rules, method = "two-key plot")
```

now how about a network graph?
```{r}
plot(topRules, method="graph")
```
Now let's see an interactive map:
```{r}
plot(topRules, method="graph", engine = 'interactive')
```

```{r}
plot(topRules, method = "grouped")
```

```{r}
plot(topRules, method = "graph",  engine = "htmlwidget")
```
now a matrix plot
```{r}
plot(topRules, method = "matrix", engine = "3d", measure = "lift")
```

moving back a bit, let's check a different set of rules:

```{r}
rules_b <- apriori(tr, parameter = list(supp=0.01, conf=1.0))
rules_b <- sort(rules_b, by='confidence', decreasing = TRUE)
summary(rules_b)
```
these more stringent criteria mean that we're down to 115 rules to sort through.
```{r}
inspect(rules_b[1:10])
```

try to look for only for rules associated with Carriage

```{r}
pneumo.rules<-sort(subset(rules_b, subset = rhs %in% "CARRIAGE"))

inspect(pneumo.rules[1:10])
```
now let's plot the carriage rules:
```{r}
plot(pneumo.rules, measure = c("support", "confidence"), shading = "lift", jitter=0)
```
```{r}
plot(pneumo.rules[1:10], method="graph", engine = 'interactive')

plot(pneumo.rules[1:10], method= "paracoord", control=list(reorder=TRUE))
```
reference: [R and Data Mining](http://www.rdatamining.com/examples/association-rules)
there are other cool market basket analysis visualizations here: 
