---
title: "exploringNepal"
format: pdf
editor: visual
---

```{r}
# Loading in required libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(MASS)
```

Making a bar chart of the serotypes in each of the four vaccines

```{r}
# Reading in the csv
nepal2Df = read.csv("nepal_gps.csv")
nepal2Df
```

```{r}
# Loading in the typhoid data set used for offset
typhoidDf = read.csv("typhoidMetadata.csv")
typhoidDf
```

```{r}
# Defining the custom color palette
custom_colors <- c(
  "PCV10" = "#00356B",   # Yale Blue
  "PCV13" = "#6A0DAD",   # Purple
  "PCV15" = "#008080",   # Teal
  "PCV20" = "#E69F00",   # Gold
  "Other" = "#999999"    # Grey fallback
)
```

```{r}
# Renaming the year column for easy merging
names(typhoidDf)[names(typhoidDf) == "year"] <- "Year"
```

```{r}
# Creating a new column that is the counts of the cases for each year
t2 <- typhoidDf %>%
  # Filtering out the years we don't want
  filter(Year > 2008 & Year < 2019) %>%
  group_by(Year) %>%
  dplyr::summarize(typhoidCount = n())

# Making sure we only have one count for each year in the dataframe
#t2 <- typhoidDf_grouped %>%
#   group_by(Year, typhoidCount)

# Creating a new column that is the counts of cases in each year
nepal2DfGrouped <- nepal2Df %>%
  group_by(Year) %>%
  dplyr::summarize(IPDCount = n()) 

# Making sure we only have one count for each year in the dataframe
#nepal2DfSelected <- nepal2DfGrouped %>%    
#   group_by(Year, IPDCount)

# Adding the typhoid counts to the nepal IPD dataset
mergedDF <- merge(nepal2DfGrouped, t2, by = "Year")
```

```{r}
# Selecting the only columns we need
nepalForBarChart <- nepal2Df %>%
  dplyr::select(In_silico_serotype, PCV7, PCV13, PCV15, PCV20)
```

```{r}
# Creating bar charts for each of the four vaccines based on the serotypes they have
ggplot(nepalForBarChart, aes(x = PCV7, fill = In_silico_serotype)) +
  geom_bar() +                    
  labs(y = "Frequency", x = "Response") 

ggplot(nepalForBarChart, aes(x = PCV13, fill = In_silico_serotype)) +
  geom_bar() +                    
  labs(y = "Frequency", x = "Response") 


ggplot(nepalForBarChart, aes(x = PCV15, fill = In_silico_serotype)) +
  geom_bar() +                    
  labs(y = "Frequency", x = "Response") +
  theme_minimal()

ggplot(nepalForBarChart, aes(x = PCV20, fill = In_silico_serotype)) +
  geom_bar() +                    
  labs(y = "Frequency", x = "Response") +
  theme_minimal()
```

```         
# Identifying the unique serotypes in the dataset
unique(nepal2Df$In_silico_serotype
```

```{r}
# Grouping all the pcv10 serotypes into the PCV10_group category
pcv10_serotypes <- c('1', '4', '5', '6B', '7F', '9V', '14', '18C', '19F', '23F')

nepalForBarChart$PCV10_group <- ifelse(nepalForBarChart$In_silico_serotype %in% pcv10_serotypes,"PCV10", "Non-PCV10")
```

```{r}
# Creating a new dataframe for each vaccine that has a more organized layout for analysis
pcv7df <- data.frame(
  Vaccine = "PCV7",
  Response = nepalForBarChart$PCV7,
  PCV10_group = nepalForBarChart$PCV10_group,
  Serotype = nepalForBarChart$In_silico_serotype)

pcv13df <- data.frame(
  Vaccine = "PCV13",
  Response = nepalForBarChart$PCV13,
  PCV10_group = nepalForBarChart$PCV10_group,
  Serotype = nepalForBarChart$In_silico_serotype)

pcv15df <- data.frame(
  Vaccine = "PCV15",
  Response = nepalForBarChart$PCV15,
  PCV10_group = nepalForBarChart$PCV10_group,
  Serotype = nepalForBarChart$In_silico_serotype)

pcv20df <- data.frame(
  Vaccine = "PCV20",
  Response = nepalForBarChart$PCV20,
  PCV10_group = nepalForBarChart$PCV10_group,
  Serotype = nepalForBarChart$In_silico_serotype)

# Combining all four into one dataframe
combined <- rbind(pcv7df, pcv13df, pcv15df, pcv20df)
```

```{r}
## Need to filter out the four random categories above
# Creating four bar charts of each vaccine and their serotypes that are PCV10
ggplot(combined, aes(x = Response, fill = PCV10_group)) +
  geom_bar() +
  facet_grid(. ~ Vaccine) +
  labs(x = "Serotype Response", y = "Frequency", fill = "PCV10 Group") +
  theme_minimal()
```

```{r}
# Finding the unique column names
colnames(nepal2Df)
```

```{r}
# I want to be able to grey out the PCV10 serotypes but idk how
ggplot(combined, aes(x = Response, fill = Serotype)) +
  geom_bar() +
  facet_grid(. ~ Vaccine) +
  labs(x = "Serotype", y = "Frequency", fill = "Serotype") +
  theme_minimal()
```

```{r}
# Creating an intro year for when the vaccine was introducted
vax.intro.date <- as.Date('2015-01-01')
vax.intro.year <- format(vax.intro.date, "%Y")

# Creating an evaluation year for when vaccine efficacy evaluations started
vax.eval.date <- as.Date('2016-01-01')
vax.eval.year <- format(vax.eval.date, "%Y")
```

```{r}
# Creating an index variable to have an ordered, discrete variable for modeling
nepalPivotedDf <- tibble::rownames_to_column(mergedDF, var = "index") %>%
  mutate(index = as.numeric(index))
```

```{r}
# Attempting method one of time series
# Interrupted time series with disconnected segments

# Creating the 'dummy' variables that are 0 before introduction of vaccine and 1 during vaccine efficacy evaluation stage
nepalPivotedDf <- nepalPivotedDf %>%
  mutate(
    period1 = ifelse(Year >= vax.intro.year, 1, 0),
    period2 = ifelse(Year >= vax.eval.year, 1, 0)
  )

# Viewing the changes
nepalPivotedDf[, c("index", "Year", "period1", "period2")]
```

```{r}
# Filtering out the year 2014
nepalPivotedDf <- nepalPivotedDf %>%
  filter(Year != 2014)

# Creating the offsets by selecting all tyhpoid cases
nepalPivotedDf$log.offset <- log(nepalPivotedDf$typhoidCount)
```

```{r}
# Should year be here
# Using the log.offset and index to predict the count of IPD infections
nepalPivotedDf$Factored_Year = as.factor(nepalPivotedDf$Year)
nepalPivotedDf$Factored_Year = droplevels(nepalPivotedDf$Factored_Year)
table(nepalPivotedDf$Factored_Year)
dim(nepalPivotedDf[nepalPivotedDf$Year == 2009,])
dim(nepalPivotedDf[nepalPivotedDf$Year == 2011,])
dim(nepalPivotedDf[nepalPivotedDf$Year == 2012,])
dim(nepalPivotedDf[nepalPivotedDf$Year == 2013,])
dim(nepalPivotedDf[nepalPivotedDf$Year == 2015,])
dim(nepalPivotedDf[nepalPivotedDf$Year == 2016,])
dim(nepalPivotedDf[nepalPivotedDf$Year == 2017,])
dim(nepalPivotedDf[nepalPivotedDf$Year == 2018,])

factors = c(rep(0,7072),rep(1,770),rep(2,3456), rep(3,272),rep(4,1927),rep(5,3456),rep(6,45326),rep(7,105472))
factors = as.factor(factors)
nepalPivotedDf$factors = factors
#mod3 <- glm.nb(as.numeric(IPDCount) ~ index + factors + offset(log.offset)+
#          # Finding the change in disease during administration and evaluation period
#          period1  + period2, data = nepalPivotedDf)

#summary(mod3)
```

```{r}
# Plotting an infections scatterplot
p1 <- 
  ggplot(nepalPivotedDf, aes(x = Year, y = IPDCount)) +
      geom_point(color = "#00356B") +
      labs(title = "Pneomoccocal Infections Scatter Plot",
         x = "Year", y = "Count") +
      theme_linedraw() +
      theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 90))

p1

# group
```

```{r}
'''
# Making predictions with confidence intervals
pred4 <- predict(mod3, type = "response", se.fit = TRUE)

# Adding the model predictions and 95% COI to the dataframe
nepalPivotedDf.pred.its <- nepalPivotedDf %>%
  mutate(se.fit = pred4$se.fit, pred = pred4$fit) %>%
  mutate(
    conf.low = pred - 1.96 * se.fit,
    conf.high = pred + 1.96 * se.fit)

# Plotting the newly created model fitting
p4 <- p1 +
  # Adding the fitting line
  geom_line(data = nepalPivotedDf.pred.its, aes(x = Year, y = pred),
            color = "#00356B") +
  # Adding the confidence interval
  geom_ribbon(data = nepalPivotedDf.pred.its, aes(ymin = conf.low, ymax = conf.high), 
              alpha = 0.2, fill = "#008080") +
  # Changeing the title name
  labs(title = "IPD Infections with a Negative Binomial Fit\nInterupted Time Series")

p4
'''
```

```{r}
'''
# Initializing the vacccine effect variables by setting them to 0
nepalPivotedDf.counterfactual <- nepalPivotedDf
nepalPivotedDf.counterfactual$period1 <- 0
nepalPivotedDf.counterfactual$period2 <- 0

# Generating the fitted values
nepalPivotedDf.pred.its$pred4.cf <- predict(mod3, type = "response", 
  # Extracting fitted value and add to df.pred.
  newdata = nepalPivotedDf.counterfactual)

# Overlaying onto plot
p4 +
 geom_line(data = nepalPivotedDf.pred.its, aes(x = Year, y = pred4.cf),
           color='#E69F00', lty = 2)
'''
```

```{r}
'''
# Generating the rate ratio between the fitted and counterfactual values
nepalPivotedDf.pred.its$rr.its <- nepalPivotedDf.pred.its$pred/nepalPivotedDf.pred.its$pred4.cf

# Plotting the rate ratio
p6 <- 
  ggplot(nepalPivotedDf.pred.its, aes(x = Year, y = rr.its)) +
      geom_line() +
      labs(title = "Rate Ratio for the ITS Mode",
         x = "Year", y = "Rate ratio") +
      theme_linedraw() +
      theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 90)) +
      geom_hline(yintercept = 1, col = "#999999", lty = 2)

p6
'''
```

```{r}
'''
# Should year be here
# A slightly more complicated model where the slope of the line is also allowed to vary in the post vaccine period
mod3a <- glm.nb(IPDCount~index + offset(log.offset) +
                 # Adding the counterfactuals with an interaction term.
                 period1 + period1*index +
                 period2 + period2*index, data=nepalPivotedDf)

summary(mod3a)
AIC(mod3, mod3a)
'''
```

# Time Series Method 2

```{r}
intro.index <- which(nepalPivotedDf$Year == year(vax.intro.date))
eval.index <- which(nepalPivotedDf$Year == year(vax.eval.date))

nepalPivotedDf <- nepalPivotedDf %>%
  mutate(
    spl1 = ifelse(index - intro.index + 1 < 0, 0, index - intro.index + 1),
    spl2 = ifelse(index - eval.index + 1 < 0, 0, index - eval.index + 1)
  )

# Inspect the changes.
nepalPivotedDf[c("Year", "index", "spl1", "spl2")]
```

```{r}
# Should year be here
mod4 <- glm.nb(IPDCount ~ index + Year + offset(log.offset) +
                 # Post-vaccine changes.
                 spl1 + spl2, data = nepalPivotedDf)

summary(mod4)
```

```{r}
# Add the prediction using the smoothed model.
nepalPivotedDf.pred.spl <- nepalPivotedDf %>%
  mutate(pred.spl = predict(mod4, type = "response"))

# Initialize the vacccine effect variables by setting them to 0.
nepalPivotedDf.counterfactual.spl <- nepalPivotedDf
nepalPivotedDf.counterfactual.spl$spl1 <- 0
nepalPivotedDf.counterfactual.spl$spl2 <- 0

# Generate the fitted values.
nepalPivotedDf.pred.spl$pred.spl.cf <- predict(mod4, type = "response", 
  # Extract fitted value and add to df.pred.
  newdata = nepalPivotedDf.counterfactual.spl)
```

```{r}
# Make predictions with confidence intervals.
pred7 <- predict(mod4, type = "response", se.fit = TRUE)

# Add the model predictions and 95% COI to the dataframe.
nepalPivotedDf.pred.spl <- nepalPivotedDf.pred.spl %>%
  mutate(se.fit = pred7$se.fit, pred = pred7$fit) %>%
  mutate(
    conf.low = pred - 1.96 * se.fit,
    conf.high = pred + 1.96 * se.fit)

p7 <- 
  ggplot(nepalPivotedDf.pred.spl, aes(x = Year, y = IPDCount)) +
      geom_point() +
      # Add the fitted line.
      geom_line(data = nepalPivotedDf.pred.spl, aes(x = Year, y = pred.spl),
                color = "#00356B") +
      # Add the confidence interval.
      geom_ribbon(data = nepalPivotedDf.pred.spl, aes(ymin = conf.low, ymax = conf.high),
                  alpha = 0.2, fill = "#008080") +
      # Add the counterfactual line.
      geom_line(data = nepalPivotedDf.pred.spl, aes(x = Year, y = pred.spl.cf),
                color = "#E69F00", lty = 2) +
      labs(title = "IPD Infections with a Negative Binomial Fit\nITS with Spline Smoothing",
         x = "Year", y = "Counts") +
      theme_linedraw() +
      # Specify aspects of the theme plot formatting settings.
      theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 90))

p7
```

```{r}
# Generate the rate ratio between the fitted and counterfactual values.
nepalPivotedDf.pred.spl$rr.spline <- nepalPivotedDf.pred.spl$pred.spl/nepalPivotedDf.pred.spl$pred.spl.cf

p8 <- p6 +
     geom_line(data = nepalPivotedDf.pred.spl, aes(x = Year, y = rr.spline),
               color = "#E69F00", lty = 2)

p8
```

Now doing method one and two but with combined carriage data

```{r}
nepalNonGps <- read.csv("nepal.csv")
nepalNonGps
```

```{r}
# Pivoting the non gps dataset
nepalNonGpsCarriage <- nepalNonGps %>%
  # Selecting year columns
  dplyr::select(X2014n, X2015n, X2017n, X2018n, X2019n, X2021n) %>%
  # Formating data frame
  pivot_longer(cols = everything(),
               names_to = "Year",
               values_to = "Count") %>%
  # Removing X and n from the year counts
  mutate(Year = gsub("X|n", "", Year)) %>%
  group_by(Year) %>%
  slice_tail()

nepalNonGpsCarriage
```

```{r}
# Plotting the counts of carriage infections over time for non gps dataset
ggplot(data = nepalNonGpsCarriage, aes(x = as.numeric(Year), y = Count)) +
  geom_line(color = "#00356B") +
  theme_bw() +
  labs(x = "Year", y = "Count", title = "Pneumococcal Carriage Infection Counts Over Time")
```

```{r}
nepalGroupedCarriage <- nepal2Df %>%
  filter(Clinical_manifestation == "CARRIAGE") %>%
  group_by(Year) %>%
  dplyr::summarize(IPDCount = n())
```

```{r}
# Combining the two datasets
mergedCarriageDf <- merge(nepalNonGpsCarriage, nepalGroupedCarriage, by = "Year", all = TRUE)
mergedCarriageDf
```

```{r}
# Adding the total counts for carriage
mergedCarriageDf$CCount <- rowSums(mergedCarriageDf[, c("Count", "IPDCount")], na.rm = TRUE)
mergedCarriageDf
```

```{r}
# For modeling, we need to use an ordered, discrete variable. Simply, we 
# can use the rownames for this purpose.
carriage2 <- tibble::rownames_to_column(mergedCarriageDf, var = "index") %>%
  mutate(index = as.numeric(index))
```

```{r}
# Creating the 'dummy' variables that are 0 before introduction of vaccine and 1 during vaccine efficacy evaluation stage

carriage2 <- carriage2 %>%
  mutate(
    period1 = ifelse(Year >= vax.intro.year, 1, 0),
    period2 = ifelse(Year >= vax.eval.year, 1, 0)
  )

# Viewing the changes
carriage2[, c("index", "Year", "period1", "period2")]
```

```{r}
t2 <- typhoidDf %>%
  group_by(Year) %>%
  dplyr::summarize(typhoidCount = n())

#t2 <- typhoidDf_grouped %>%
#   group_by(Year, typhoidCount) %>%
#  summarize(.groups = 'drop')

#nepalGroupedCarriage <- nepal2Df %>%
#  filter(Clinical_manifestation == "CARRIAGE") %>%
#  group_by(Year) %>%
#  dplyr::summarize(IPDCount = n())

# Adding the typhoid counts to the carriage2 dataset
mergedCarriage2DF <- merge(carriage2, t2, by = "Year")
```

NOW ALL THREE METHODS WITH ONLY CARRIAGE DATA

Method One

```{r}
# Creating the offsets by selecting all tyhpoid cases
mergedCarriage2DF$log.offset <- log(mergedCarriage2DF$typhoidCount)
```

```{r}
# Making year into a factored variable
mergedCarriage2DF$Int_Year = as.integer(mergedCarriage2DF$Year)

mergedCarriage2DF <- mergedCarriage2DF %>%
  filter(Year != 2014)

mod3 <- glm.nb(CCount ~ index  + Int_Year + offset(log.offset) +
          # Finding the change in disease during administration and evaluation period
          period1  + period2, data = mergedCarriage2DF, control = glm.control(maxit = 100, epsilon = 1e-08))

summary(mod3)
```

```{r}
# Adding a new column for plotting type
#mergedCarriage2DF.pred.its$LineType <- "Fitted"
#mergedCarriage2DF.pred.its$Counterfactual <- "Counterfactual"

p1 <- 
  ggplot(mergedCarriage2DF, aes(x = Year, y = CCount)) +
      geom_point(color = "#6A0DAD") +
      labs(title = "Carriage Infections Scatter Plot",
         x = "Year", y = "Counts") +
      # Have y-axis for the two plots be the same.
      #ylim(0, NA) +
      theme_linedraw() +
      # Specify aspects of the theme plot formatting settings.
      theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 0, size = 12),
            axis.text.y = element_text(size = 12),
            axis.title = element_text(size = 14),
            title = element_text(size = 18))

p1

```

```{r}
# Make predictions with confidence intervals.
pred4 <- predict(mod3, type = "response", se.fit = TRUE)

# Add the model predictions and 95% COI to the dataframe.
mergedCarriage2DF.pred.its <- mergedCarriage2DF %>%
  mutate(se.fit = pred4$se.fit, pred = pred4$fit) %>%
  mutate(
    conf.low = pred - 1.96 * se.fit,
    conf.high = pred + 1.96 * se.fit
  )

# Plot the newly created model fitting.
p4 <- p1 +
  # Add the fitting line.
  geom_line(data = mergedCarriage2DF.pred.its, aes(x = Year, y = pred, color = "Negative Binomial Fitted Line"), size = .7, group = 1) +
  # Add the confidence interval.
  geom_ribbon(data = mergedCarriage2DF.pred.its, aes( ymin = conf.low, ymax = conf.high), fill = "#008080", group = 1, alpha = 0.2) +
  # Change the title name.
  labs(title = "Pneumococcal Carriage Infections in Nepal",
       subtitle = "Interrupted Time Series, Negative Binomial Fit")
p4
```

```{r}
# Initialize the vacccine effect variables by setting them to 0.
mergedCarriage2DF.counterfactual <- mergedCarriage2DF
mergedCarriage2DF.counterfactual$period1 <- 0
mergedCarriage2DF.counterfactual$period2 <- 0

# Generate the fitted values.
mergedCarriage2DF.pred.its$pred4.cf <- predict(mod3, type = "response", 
  # Extract fitted value and add to df.pred.
  newdata = mergedCarriage2DF.counterfactual)

# Overlay onto plot.
pFinal <- p4 +
  geom_line(data = mergedCarriage2DF.pred.its, aes(x = Year, y = pred4.cf, color = "Counterfactual"), size = .8, lty = 2, group = 1) +
      geom_vline(xintercept = 2015, color = "red", linetype = "dashed", size = .8) +

  scale_color_manual(name='Legend',
                     breaks=c('Negative Binomial Fitted Line', 'Counterfactual'),
                     values=c("Negative Binomial Fitted Line" = "#00356B","Counterfactual" = "#E69F00")) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 12),
        legend.title = element_blank())


pFinal
```

```{r}
# Generate the rate ratio between the fitted and counterfactual values.
mergedCarriage2DF.pred.its$rr.its <- mergedCarriage2DF.pred.its$pred/mergedCarriage2DF.pred.its$pred4.cf

p6 <- 
  ggplot(mergedCarriage2DF.pred.its, aes(x = Year, y = rr.its)) +
      geom_line(group = 1) +
      labs(title = "Rate Ratio Between Model and Counterfactual",
         x = "Year", y = "Rate ratio") +
      theme_linedraw() +
      # Specify aspects of the theme plot formatting settings.
      theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 90)) +
      geom_hline(yintercept = 1, col = "#999999", lty = 2)

p6
# look up rate ratio comparison
```

```{r}
# Adding in empty rows in the dataset so year goes up by 1 each time
#emptyYears <- data.frame(Year  = c(2010, 2011, 2013, 2016))
#mergedTest <- merge(mergedCarriage2DF, emptyYears, by = "Year", all = TRUE)
```

```{r}
method3Carriage <- tibble::rownames_to_column(mergedCarriage2DF, var = "index3") %>%
 mutate(index3 = as.numeric(index3))
```

```{r}
# A slightly more complicated model where the slope of the line is also allowed to vary in the post vaccine period
mod3a <- glm.nb(CCount ~ index3 + offset(log.offset) +
                 # Add the counterfactuals with an interaction term.
                 period1 + period1*index3 +
                 period2 + period2*index3, data = method3Carriage)
 
summary(mod3a)
AIC(mod3, mod3a)
```

Method Two

```{r}
# Changing the eval date because there is no 2016
vax.eval.date <- as.Date('2017-01-01')
vax.eval.year <- format(vax.eval.date, "%Y")

intro.index <- which(method3Carriage$Year == year(vax.intro.date))
eval.index <- which(method3Carriage$Year == year(vax.eval.date))

intro.index <- as.numeric(intro.index)
eval.index <- as.numeric(eval.index)

method3Carriage <- method3Carriage %>%
  mutate(
    spl1 = ifelse(index3 - intro.index + 1 < 0, 0, index3 - intro.index + 1),
    spl2 = ifelse(index3 - eval.index + 1 < 0, 0, index3  - eval.index + 1)
  )

# Inspect the changes.
method3Carriage[c("Year", "index", "spl1", "spl2")]
```

```{r}
method3Carriage <- method3Carriage %>%
  filter(Year != 2014)
mod4 <- glm.nb(CCount ~ index3  + offset(log.offset) +
                 # Post-vaccine changes.
                 spl1 + spl2, data = method3Carriage)

summary(mod4)
```

```{r}
# Add the prediction using the smoothed model.
method3Carriage.pred.spl <- method3Carriage %>%
  mutate(pred.spl = predict(mod4, type = "response"))

# Initialize the vacccine effect variables by setting them to 0.
method3Carriage.counterfactual.spl <- method3Carriage
method3Carriage.counterfactual.spl$spl1 <- 0
method3Carriage.counterfactual.spl$spl2 <- 0

# Generate the fitted values.
method3Carriage.pred.spl$pred.spl.cf <- predict(mod4, type = "response", 
  # Extract fitted value and add to df.pred.
  newdata = method3Carriage.counterfactual.spl)
```

```{r}
# Make predictions with confidence intervals.
pred7 <- predict(mod4, type = "response", se.fit = TRUE)

# Add the model predictions and 95% COI to the dataframe.
method3Carriage.pred.spl <- method3Carriage.pred.spl %>%
  mutate(se.fit = pred7$se.fit, pred = pred7$fit) %>%
  mutate(
    conf.low = pred - 1.96 * se.fit,
    conf.high = pred + 1.96 * se.fit
  )


p7 <- 
  ggplot(method3Carriage.pred.spl, aes(x = Year, y = CCount)) +
      geom_point() +
      # Add the fitted line.
      geom_line(data = method3Carriage.pred.spl, aes(x = Year, y = pred.spl, color = "Negative Binomial Fitted Line"), size = .7, group = 1) +
      # Add the confidence interval.
      geom_ribbon(data = method3Carriage.pred.spl, aes(ymin = conf.low, ymax = conf.high), fill = "#008080",
                  alpha = 0.2, group = 1) +
      # Add the counterfactual line.
      geom_line(data = method3Carriage.pred.spl, aes(x = Year, y = pred.spl.cf, color = "Counterfactual"), size = .8,
         lty = 2, group = 1) +
      labs(title = "Pneumoccocal Carriage Infections in Nepal",
      subtitle = "Negative Binomial Fit,\nInterrupted Time Series with Spline Smoothing",
         x = "Year", y = "Counts") +
      theme_linedraw() +
    geom_vline(xintercept = 2015, color = "red", linetype = "dashed", size = .8) +

    scale_color_manual(name='Legend',
                     breaks=c('Negative Binomial Fitted Line', 'Counterfactual'),
                     values=c("Negative Binomial Fitted Line" = "#00356B","Counterfactual" = "#E69F00")) +
      # Specify aspects of the theme plot formatting settings.
      theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 0, size = 12),
            axis.text.y = element_text(size = 12),
            axis.title = element_text(size = 12),
            axis.title.x = element_text(size = 14),
            axis.title.y =element_text(size = 14),
            title = element_text(size = 18),
            legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 12)) +
      scale_x_continuous(breaks = seq(floor(min(method3Carriage.pred.spl$Year)), 
                                  ceiling(max(method3Carriage.pred.spl$Year)), 
                                  by = 1)) +
                                 annotate("text", x = 2015.15, y = 00.1, label = "Vaccine Introduced", color = "red", hjust = 0)

p7
# use subtitle
# angles years
```

```{r}
# Generate the rate ratio between the fitted and counterfactual values.
method3Carriage.pred.spl$rr.spline <- method3Carriage.pred.spl$pred.spl/method3Carriage.pred.spl$pred.spl.cf

p8 <- p6 +
     geom_line(data = method3Carriage.pred.spl, aes(x = Year, y = rr.spline),
               color = "#E69F00", lty = 2) + 
      # Update the title.
      labs(title = "Rate Ratio with the Spline Model")

p8
```

typhoid dataset ends at 2019, so we left out 2021 from the nepal datasets

Method three

```{r}
# Setting the count variable to NA during the post vaccine period

method3Carriage$CCountPre <- method3Carriage$CCount

introYear <- year(vax.intro.date)
method3Carriage$Year <- as.numeric(method3Carriage$Year)



method3Carriage$CCountPre[which(method3Carriage$Year >= 2016)] <- NA

method3Carriage[, c("Year", "CCountPre")]
```

```{r}

# Making index into an integer
#method3Carriage$int_index <- as.numeric(method3Carriage$index)

#pre_data <- method3Carriage %>% filter(!is.na(CCountPre))

#mod5 <- glm.nb(CCountPre ~ int_index + offset(log.offset), data = pre_data)


# Not adding year because no seasonality
mod5<- glm.nb(CCountPre ~ index + offset(log.offset), data = method3Carriage)

# Add the prediction using the smoothed model.
method3Carriage.pred.pre <- method3Carriage %>%
  mutate(pred.mod5 = predict(mod5, type = "response", newdata = method3Carriage))
```

```{r}
## Prepare a plot to visualize the factual model predictions

# Make predictions with confidence intervals.
pred7 <- predict(mod5, type = "response", se.fit = TRUE, newdata = method3Carriage)

# Add the model predictions and 95% COI to the dataframe.
method3Carriage.pred.pre <- method3Carriage.pred.pre %>%
  mutate(se.fit = pred7$se.fit, pred = pred7$fit) %>%
  mutate(
    conf.low = pred - 1.96 * se.fit,
    conf.high = pred + 1.96 * se.fit
  )


p_m3 <- 
  ggplot(method3Carriage.pred.pre, aes(x = Year, y = CCount)) +
      geom_point() +
      # Add the fitted line.
      geom_line(data = method3Carriage.pred.pre, aes(x = Year, y = pred),
                color = "#00356B") +
      # Add the confidence interval.
      geom_ribbon(data = method3Carriage.pred.pre, aes(ymin = conf.low, ymax = conf.high),
                  alpha = 0.2, fill = "#008080") +
      labs(title = "Pneumococcal Carriage Infections",
           subtitle = "Negative Binomial Fit,\nExtrapolating Based on the Pre-Vaccine Period",
         x = "Year", y = "Counts") +
      theme_linedraw() +
    scale_x_continuous(breaks = seq(floor(min(method3Carriage.pred.pre$Year)), 
                                  ceiling(max(method3Carriage.pred.pre$Year)), 
                                  by = 1)) +
        theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 0, size = 12),
            axis.text.y = element_text(size = 12),
            axis.title = element_text(size = 12),
            axis.title.x = element_text(size = 14),
            axis.title.y =element_text(size = 14),
            title = element_text(size = 18))
p_m3
```

```{r}
ds.pred.mod5$rr.trend<-method3Carriage$CCount/ method3Carriage.pred.mod5$pred.mod5
```

```{r}
# Change color here
p8 +
  geom_line(color='#377eb8',lwd=1, lty=1,data=ds.pred.mod5, aes(x=Year, y=rr.trend))
```

Comparing the three methods

```{r}
eval.period<-method3Carriage$Year> vax.eval.year

rr.its.eval<- mean(mergedCarriage2DF.pred.its$rr.its[eval.period])

rr.spline.eval<- mean(method3Carriage.pred.spl$rr.spline[eval.period])

rr.trend.eval<- sum(ds.pred.mod5$CCount[eval.period])/sum(ds.pred.mod5$pred.mod5[eval.period])

round(c('ITS'=rr.its.eval,'Spline'=rr.spline.eval,'Trend Extrapolation'=rr.trend.eval ),2)

#Percent decline
100*(1- round(c('ITS'=rr.its.eval,'Spline'=rr.spline.eval,'Trend Extrapolation'=rr.trend.eval ),2))
```

One of the modeling methods identified a decrease of approximately 73% in pneumonia deaths due to pneumococcal disease after the release of the PCV10 vaccine and one identified a 99% decrease. The extrapolation method indicates a weaker vaccine effect; however, we expect some of this result to be spurious due to the level of random noise introduced by applying the observed values in the counterfactual ratio calculation.

```{r}
mergedCarriage2DF.pred.its$rr.its.a <- mergedCarriage2DF.pred.its$pred/mergedCarriage2DF.pred.its$pred4.cf.a

```

```{r}
p_ratio <- 
  ggplot(df.pred.its, aes(x = date, y = rr.its.a)) +
      geom_line() +
      labs(title = "Rate Ratio of the ITS Model",
         x = "Date", y = "Rate ratio") +
      ylim(0, NA) +
      geom_vline(xintercept = vax.intro.date, col = "red", lty = 2) +
      annotate("text", x = vax.intro.date + 30, y = 00.1, label = "Vaccine Introduced", color = "red", hjust = 0) +
      theme_linedraw()

p_ratio +
  geom_line(data = df.pred.its, aes(x = date, y = rr.its.b),
            color = "#4daf4a")
```

```{r}
# Inspect methods sensitivity to vaccine impact.
p_ratio <- 
  ggplot(df.pred.its, aes(x = date, y = rr.its.a)) +
      geom_line() +
      labs(title = "Rate Ratio of the ITS Model",
         x = "Date", y = "Rate ratio") +
      ylim(0, NA) +
      geom_vline(xintercept = vax.intro.date, col = "red", lty = 2) +
      annotate("text", x = vax.intro.date + 30, y = 00.1, label = "Vaccine Introduced", color = "red", hjust = 0) +
      theme_linedraw()

p_ratio +
  geom_line(data = df.pred.its, aes(x = date, y = rr.its.b),
            color = "#4daf4a")
```

```{r}
p6 <- 
  ggplot(mergedCarriage2DF.pred.its, aes(x = Year, y = rr.its)) +
      geom_line(group = 1) +
      labs(title = "Rate Ratio Between Model and Counterfactual",
         x = "Year", y = "Rate ratio") +
      theme_linedraw() +
      # Specify aspects of the theme plot formatting settings.
      theme(panel.spacing = unit(2, 'lines'), 
            axis.text.x = element_text(angle = 90)) +
      geom_hline(yintercept = 1, col = "#999999", lty = 2) + 
     geom_line(data = method3Carriage.pred.spl, aes(x = Year, y = rr.spline),
               color = "#E69F00", lty = 2) + 
      # Update the title.
      labs(title = "Rate Ratio with the Spline Model")

p6
```

```{r}
mergedCarriage2DF.pred.its$Year <- as.numeric(mergedCarriage2DF.pred.its$Year)
method3Carriage.pred.spl$Year <- as.numeric(method3Carriage.pred.spl$Year)

p6 <- ggplot() +
  geom_line(data = mergedCarriage2DF.pred.its, 
            aes(x = Year, y = rr.its, color = "Interrupted Time Series Model", group = 1), 
            linetype = "solid", size = 1.2) +
  geom_line(data = method3Carriage.pred.spl, 
            aes(x = Year, y = rr.spline, color = "Spline Model", group = 1), size = 1.2) +
  geom_hline(yintercept = 1, color = "grey20", lty = 2) +
  
  geom_vline(xintercept = 2015, color = "red", linetype = "dashed", size = .8) +
  
  labs(title = "Rate Ratios of Interrupted Time Series and Spline Model",
       x = "Year", y = "Rate Ratio", color = "Model") +
  theme_linedraw() +
  theme(panel.spacing = unit(2, 'lines'), 
        axis.text.x = element_text(angle = 90)) +
  scale_color_manual(values = c("Interrupted Time Series Model" = "#6A0DAD", "Spline Model" = "#008080")) +
  scale_x_continuous(breaks = seq(floor(min(mergedCarriage2DF.pred.its$Year)), 
                                  ceiling(max(mergedCarriage2DF.pred.its$Year)), 
                                  by = 1)) +
  theme(panel.spacing = unit(2, 'lines'),
        legend.position = "bottom",
        legend.title = element_blank(),
         axis.text.x = element_text(angle = 0, size = 12),
            axis.text.y = element_text(size = 12),
            axis.title = element_text(size = 12),
            axis.title.x = element_text(size = 14),
            axis.title.y =element_text(size = 14),
            title = element_text(size = 15),
        legend.text = element_text(size = 12)) +
        annotate("text", x = 2015.15, y = 00.1, label = "Vaccine Introduced", color = "red", hjust = 0)


p6
```

Trying method three again

```{r}
# Create a new variable where it is J12_J18_prim until the vaccine is introduced,
# then it is NA afterwards.

method3Carriage$CCountPre <- method3Carriage$CCount

method3Carriage$CCountPre[which(method3Carriage$Year >= 2016)] <- NA

# Inspect the changes.
method3Carriage[sort(sample(1:6, 6)), c("Year", "CCountPre")]
```

```{r}
mod_method3 <- glm.nb(CCountPre ~ index + offset(log.offset), data = method3Carriage)

# Add the prediction using the smoothed model.
method3Carriage.pred.pre <- method3Carriage %>%
  mutate(pred.pre = predict(mod_method3, type = "response", newdata = method3Carriage))
```

```{r}
method3Carriage.pred.pre$rr.trend <- method3Carriage$CCount/method3Carriage.pred.pre$pred.pre

```

```{r}
## --------------------
## Prepare a plot to visualize the factual model predictions

# Make predictions with confidence intervals.
pred7 <- predict(mod_method3, type = "response", se.fit = TRUE, newdata = method3Carriage)

# Add the model predictions and 95% COI to the dataframe.
method3Carriage.pred.pre <- method3Carriage.pred.pre %>%
  mutate(se.fit = pred7$se.fit, pred = pred7$fit) %>%
  mutate(
    conf.low = pred - 1.96 * se.fit,
    conf.high = pred + 1.96 * se.fit
  )


p_m3 <- 
  ggplot(method3Carriage.pred.pre, aes(x = Year, y = CCount)) +
      geom_point() +
      # Add the fitted line.
      geom_line(data = method3Carriage.pred.pre, aes(x = Year, y = pred.pre),
                color = "#377eb8") +
      # Add the confidence interval.
      geom_ribbon(data = method3Carriage.pred.pre, aes(ymin = conf.low, ymax = conf.high),
                  alpha = 0.2, fill = "blue") +
      labs(title = "Deaths with a Negative Binomial Fit\nExtrapolating Based on the Pre-Vaccine Period",
         x = "Year", y = "Counts") +
      theme_linedraw()

p_m3
```

```{r}
p_ratio <- p6 +
  geom_line(data = method3Carriage.pred.pre, aes(x = Year, y = rr.trend),
            color = '#377eb8', size = 1.2) +
  # Update the title.
  labs(title = "Rate Ratio with the Extrapolation Model")

p_ratio
```

```{r}
# Create a Boolean variable that defines the vaccine evaluation period.
eval.period <- method3Carriage$Year > 2015
eval.period2 <- mergedCarriage2DF$Year > 2015

# Calculate the average outcome during the vaccine evaluation period.
rr.its.eval    <- mean(mergedCarriage2DF.pred.its$rr.its[eval.period2])
rr.spline.eval <- mean(method3Carriage.pred.spl$rr.spline[eval.period])
rr.trend.eval  <- sum(method3Carriage.pred.pre$CCount[eval.period])/sum(method3Carriage.pred.pre$pred.pre[eval.period])

# Average decline detected from 1.
round(c("ITS - Disconnected" = rr.its.eval, "ITS - Spline" = rr.spline.eval,
        "Extrapolation" = rr.trend.eval), digits = 2)
```

```{r}
# Percent decline detected.
100*(1 - round(c("ITS - Disconnected" = rr.its.eval, "ITS - Spline" = rr.spline.eval, 
                 "Extrapolation" = rr.trend.eval), digits = 2))
```

```{r}
p6 <- ggplot() +
  geom_line(data = mergedCarriage2DF.pred.its, 
            aes(x = Year, y = rr.its, color = "Interrupted Time Series Model", group = 1), 
            linetype = "solid", size = 1.2) +
  geom_line(data = method3Carriage.pred.spl, 
            aes(x = Year, y = rr.spline, color = "Spline Model", group = 1), size = 1.2) +
  geom_line(data = method3Carriage.pred.pre, aes(x = Year, y = rr.trend, color = 'Extrapolation Model', group = 1),
             size = 1.2) +
  geom_hline(yintercept = 1, color = "grey20", lty = 2) +
  
  geom_vline(xintercept = 2015, color = "red", linetype = "dashed", size = .8) +
  
  labs(title = "Rate Ratios of Interrupted Time Series, Spline,\nand Extrapolation Models",
       x = "Year", y = "Rate Ratio", color = "Model") +
  theme_linedraw() +
  theme(panel.spacing = unit(2, 'lines'), 
        axis.text.x = element_text(angle = 90)) +
  scale_color_manual(values = c("Interrupted Time Series Model" = "#6A0DAD", "Spline Model" = "#008080", "Extrapolation Model" = "#E69F00")) +
  scale_x_continuous(breaks = seq(floor(min(mergedCarriage2DF.pred.its$Year)), 
                                  ceiling(max(mergedCarriage2DF.pred.its$Year)), 
                                  by = 1)) +
  theme(panel.spacing = unit(2, 'lines'),
        legend.position = "bottom",
        legend.title = element_blank(),
         axis.text.x = element_text(angle = 0, size = 12),
            axis.text.y = element_text(size = 12),
            axis.title = element_text(size = 12),
            axis.title.x = element_text(size = 14),
            axis.title.y =element_text(size = 14),
            title = element_text(size = 15),
        legend.text = element_text(size = 12)) +
        annotate("text", x = 2015.15, y = 00.1, label = "Vaccine Introduced", color = "red", hjust = 0)

p6
```
